\bf An observational study of papers published in \emph{PLOS ONE} and studies posted to a trial registry.

### 2021-04-21 manuscript/asa_template/asa_template.Rmd
add plos boilerplate text

@@

>>>>

* Results of cosine similaritites

```{r}
load('../../results/plos.cosinesim.10topics.rda')

ftab = flextable(ftab.cosine.plos) %>% theme_box() %>% autofit()
ftab <- set_caption(ftab, "Summary of cosine similarity scores by topic: PLOS ONE")
ftab

@@

We downloaded 28,008 studies. The numbers of excluded studies are shown in Figure~X. 
Of the 12,700 included studies, 9,523 (75%) had a statistical methods section.

The median length of the section was 129 words with an inter-quartile range of 71 to 219 words.
Some methods sections were only one word, including "ANOVA", "t-test", "SPSS" and even "SSPS".

The clustering algorithm found groups that were purely sample size calculations (topic 2), pilot studies (topic 5), safety/tolerability studies (topic 6) and repeated measures ANOVA (topic 10). There were cases where the exact same method section had been re-used in a different study.

We also found evidence of 'boilerplate' sections clustered as topic 3, example text

**Box x: Examples of boilerplate text from ANZCTR**

Topic 3

* "Shapiro Wilk test was used as normality test. Continuous variables were compared using Student t-test and Mann-Whitney U test when the data were not normally distributed. Categorical variables were compared using Pearson’s chi-squared test and Fisher’s exact test. Paired data were analyzed using Paired t-test and Wilcoxon signed rank test when data were not normally distributed."
* "Comparisons between categorical variables will be made either using chi square or Fisher exact test. Continuous data will be compared using the Student’s t-test or Mann-Whitney U test. Two sided p values of less than 0.05 will be considered statistically significant."

We examined if four study characteristics were associated with a missing statistics section. The odds ratios and 95% credible intervals are in Table~X. Observational studies were less likely to have a missing methods section compared with interventional studies. Missing sections became less likely over time. Studies with more funders and a larger target sample size were less likely to have a missing methods section.

```{r}
table =  read.table(header=TRUE, sep='&', text='
Variable & Odds ratio & 95% CI
Study type = Observational & 0.78 & 0.69, 0.89
Date (per year) & 0.90 & 0.88, 0.91
Number of funders & 0.80 & 0.74, 0.86
Target sample size (per doubling) & 0.90 & 0.88, 0.92')

ftab = flextable(table) %>%
theme_box() %>%
autofit()
ftab
```

* Final sample size

# Discussion

The first line in many statistical analysis sections in _PLOS ONE_ was the software used and some entire sections in ANZCTR only stated the software, implying that the software is the most important detail. As Doug Altman said, "Many people think that all you need to do statistics is a computer and appropriate software" [@Altman1994]. This is far from the truth, and whilst it is important for researchers to mention the software and version used for reproducibility purposes, it is a minor detail compared with detailing what methods were used and why.

A frequent theme in the boilerplate statistical methods is the definition of statistical significance, nearly always using a p-value at the 5% level. This widespread use of statistical significance is troubling giving the bright-line thinking it engenders [@McShane2019] and the common misinterpretations of p-values [@Goodman2008].

Despite the extensive array of statistical tests available, many authors are reporting the same few methods.

One reason these inadequate sections get published is that most journals do not use statistical reviewers, despite empirical evidence showing they improve manuscript quality [@Hardwicke2020].

A related paper has criticised vague statistical methods sections because they deprive readers and reviewers for the opportunity to confirm that the appropriate methods were used [@Weissgerber2018]. These authors checked hundreds of papers using ANOVA and found that 95\% did not contain the information needed to determine what type of ANOVA was performed. This lack of information could well be because the authors used a boilerplate statistical methods section that was missing key details.

If authors shared their code then this would provide an alternative route for checking what statistical methods were used. This is not a perfect solution, as we still want authors to accurately report their methods, but it does increase transparency. However, a recent paper found that code sharing was very low in biomedical papers, with just 2\% of a sample of over 6,000 papers sharing code [@Serghiou2021].

Many researchers are using lazy practice by copying a standard "boilerplate" statistical methods section, likely cut-and-pasting from other researchers or projects. This is a strong sign of the ritualistic practice of statistics where researchers go through the motions rather than using conscientious practice [@Stark2018]. This is concerning because using the wrong statistical methods can reduce the value of study, or worse, invalidate the entire study. These mistakes are avoidable and are wasting of thousands of hours of researchers' time and the time of patients and volunteers. Poor statistical practice is a key driver of the ongoing reproducibility crisis in science [@Ioannidis2014].

## Limitations

We did not check whether papers used the correct methods, and for some simple studies a 'boilerplate' statistical methods might be adequate.

We examined papers where there was a statistics section, and we missed papers that used statistical analysis but did not include a statistical analysis section. Reiterate outcomes of random sample checking here.

We only examined one large journal and one trial registry and hence our results may not be generalisable to all journals or registries, especially those that consistently use a statistical reviewer.

We searched the full text of _PLOS ONE_ papers but not the supporting information which may contain statistical methods sections for some papers. The search terms we used to find statistical methods appeared in the supporting information titles for xxx papers (x%). We did not include the supporting information because it is less structured than the paper and could be in PDF or Word format.

### 2021-04-21 manuscript/asa_template/asa_template.tex
add plos boilerplate text

@@

\caption{ANZCTR topic clouds.}\label{fig:unnamed-chunk-7}
\end{figure}

@@

\caption{PLOS ONE clustering metrics.}\label{fig:unnamed-chunk-8}
\end{figure}

@@

\caption{PLOS ONE topic values.}\label{fig:unnamed-chunk-9}
\end{figure}

@@

\caption{ANZCTR clustering metrics.}\label{fig:unnamed-chunk-10}
\end{figure}

@@

\caption{ANZCTR topic values.}\label{fig:unnamed-chunk-11}
\end{figure}

>>>>

\begin{itemize}
\tightlist
\item
  Results of cosine similaritites
\end{itemize}

\includegraphics[width=6.13in,height=3.49in,keepaspectratio]{asa_template_files/figure-latex/unnamed-chunk-7-1.png}

@@

\caption{ANZCTR topic clouds.}\label{fig:unnamed-chunk-8}
\end{figure}

We downloaded 28,008 studies. The numbers of excluded studies are shown
in Figure\textasciitilde X. Of the 12,700 included studies, 9,523 (75\%)
had a statistical methods section.

The median length of the section was 129 words with an inter-quartile
range of 71 to 219 words. Some methods sections were only one word,
including ``ANOVA'', ``t-test'', ``SPSS'' and even ``SSPS''.

The clustering algorithm found groups that were purely sample size
calculations (topic 2), pilot studies (topic 5), safety/tolerability
studies (topic 6) and repeated measures ANOVA (topic 10). There were
cases where the exact same method section had been re-used in a
different study.

We also found evidence of `boilerplate' sections clustered as topic 3,
example text

\textbf{Box x: Examples of boilerplate text from ANZCTR}

Topic 3

\begin{itemize}
\tightlist
\item
  ``Shapiro Wilk test was used as normality test. Continuous variables
  were compared using Student t-test and Mann-Whitney U test when the
  data were not normally distributed. Categorical variables were
  compared using Pearson's chi-squared test and Fisher's exact test.
  Paired data were analyzed using Paired t-test and Wilcoxon signed rank
  test when data were not normally distributed.''
\item
  ``Comparisons between categorical variables will be made either using
  chi square or Fisher exact test. Continuous data will be compared
  using the Student's t-test or Mann-Whitney U test. Two sided p values
  of less than 0.05 will be considered statistically significant.''
\end{itemize}

We examined if four study characteristics were associated with a missing
statistics section. The odds ratios and 95\% credible intervals are in
Table\textasciitilde X. Observational studies were less likely to have a
missing methods section compared with interventional studies. Missing
sections became less likely over time. Studies with more funders and a
larger target sample size were less likely to have a missing methods
section.

\includegraphics[width=4.66in,height=1.55in,keepaspectratio]{asa_template_files/figure-latex/unnamed-chunk-9-1.png}

\begin{itemize}
\tightlist
\item
  Final sample size
\end{itemize}

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

The first line in many statistical analysis sections in \emph{PLOS ONE}
was the software used and some entire sections in ANZCTR only stated the
software, implying that the software is the most important detail. As
Doug Altman said, ``Many people think that all you need to do statistics
is a computer and appropriate software'' \citep{Altman1994}. This is far
from the truth, and whilst it is important for researchers to mention
the software and version used for reproducibility purposes, it is a
minor detail compared with detailing what methods were used and why.

A frequent theme in the boilerplate statistical methods is the
definition of statistical significance, nearly always using a p-value at
the 5\% level. This widespread use of statistical significance is
troubling giving the bright-line thinking it engenders
\citep{McShane2019} and the common misinterpretations of p-values
\citep{Goodman2008}.

Despite the extensive array of statistical tests available, many authors
are reporting the same few methods.

One reason these inadequate sections get published is that most journals
do not use statistical reviewers, despite empirical evidence showing
they improve manuscript quality \citep{Hardwicke2020}.

A related paper has criticised vague statistical methods sections
because they deprive readers and reviewers for the opportunity to
confirm that the appropriate methods were used \citep{Weissgerber2018}.
These authors checked hundreds of papers using ANOVA and found that 95\%
did not contain the information needed to determine what type of ANOVA
was performed. This lack of information could well be because the
authors used a boilerplate statistical methods section that was missing
key details.

If authors shared their code then this would provide an alternative
route for checking what statistical methods were used. This is not a
perfect solution, as we still want authors to accurately report their
methods, but it does increase transparency. However, a recent paper
found that code sharing was very low in biomedical papers, with just 2\%
of a sample of over 6,000 papers sharing code \citep{Serghiou2021}.

Many researchers are using lazy practice by copying a standard
``boilerplate'' statistical methods section, likely cut-and-pasting from
other researchers or projects. This is a strong sign of the ritualistic
practice of statistics where researchers go through the motions rather
than using conscientious practice \citep{Stark2018}. This is concerning
because using the wrong statistical methods can reduce the value of
study, or worse, invalidate the entire study. These mistakes are
avoidable and are wasting of thousands of hours of researchers' time and
the time of patients and volunteers. Poor statistical practice is a key
driver of the ongoing reproducibility crisis in science
\citep{Ioannidis2014}.

\hypertarget{limitations}{%
\subsection{Limitations}\label{limitations}}

We did not check whether papers used the correct methods, and for some
simple studies a `boilerplate' statistical methods might be adequate.

We examined papers where there was a statistics section, and we missed
papers that used statistical analysis but did not include a statistical
analysis section. Reiterate outcomes of random sample checking here.

We only examined one large journal and one trial registry and hence our
results may not be generalisable to all journals or registries,
especially those that consistently use a statistical reviewer.

We searched the full text of \emph{PLOS ONE} papers but not the
supporting information which may contain statistical methods sections
for some papers. The search terms we used to find statistical methods
appeared in the supporting information titles for xxx papers (x\%). We
did not include the supporting information because it is less structured
than the paper and could be in PDF or Word format.

@@

\caption{PLOS ONE clustering metrics.}\label{fig:unnamed-chunk-10}
\end{figure}

@@

\caption{PLOS ONE topic values.}\label{fig:unnamed-chunk-11}
\end{figure}

@@

\caption{ANZCTR clustering metrics.}\label{fig:unnamed-chunk-12}
\end{figure}

@@

\caption{ANZCTR topic values.}\label{fig:unnamed-chunk-13}
\end{figure}
